\Title

### 2013-02-04 supplementary.tex
PCA of good patient in suppl

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Selection of the patient data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.48\linewidth]{Shankarappa_PCA_p7}
\includegraphics[width=0.48\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7}
\caption{Left: PCA of all sequences from patient p7 (colors indicate time from seroconversion,
from blue to red). Right: Allele frequency trajectories for nonsynonymous
changes in the same patient.}
\label{fig:aftp7}
\end{center}
\end{figure}
Not all of the patients studied in refs.~\citealp{shankarappa_consistent_1999} and
\citealp{bunnik_autologous_2008} were analyzed. The reason is that patients p4,
p7, p8, p9 from ref.~\citealp{shankarappa_consistent_1999} and patients ACH19542
and ACH19768 from ref.~\citealp{bunnik_autologous_2008} showed clear signs of
compartimentalization (population structure) during infection, which would have
been an irreparable confounder for our analysis. The difference is clear if
PCA is performed on all sequences from a ``compartimentalized'' patient, p7.
This is shown in \figurename~\ref{fig:aftp7}: late sequences subdivide in two
well-defined clusters -- this characteristic is absent from all the patient
sequences used in our main analysis. On the right, allele frequency trajectories from the same
patient are plotted: evidently, no derived allele is fixed after 1000 days
from seroconversion, but a lot of mutations swarm to frequencies around 0.5-0.8. Of
course, the fixation probability of new mutations in such a sample is very low,
but this is an effect of population structure rather than any
of the processes discussed in the main text (time-dependent selection, competition
of mutants within the same epitope).

\clearpage

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Selection of the patient data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.4\linewidth]{Shankarappa_PCA_p1}
\includegraphics[width=0.4\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1}
\includegraphics[width=0.4\linewidth]{Shankarappa_PCA_p7}
\includegraphics[width=0.4\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7}
\caption{Panel A) PCA of all sequences from patient p1 (colors indicate time from seroconversion,
from blue to red). Panel B) Allele frequency trajectories for nonsynonymous
changes in the same patient. Panels C and D) the same plots from patient p7, whose
sequences show signs of compartimentalization.}
\label{fig:aftp}
\end{center}
\end{figure}
Not all of the patients studied in refs.~\citealp{shankarappa_consistent_1999} and
\citealp{bunnik_autologous_2008} were analyzed. The reason is that patients p4,
p7, p8, p9 from ref.~\citealp{shankarappa_consistent_1999} and patients ACH19542
and ACH19768 from ref.~\citealp{bunnik_autologous_2008} showed clear signs of
compartimentalization (population structure) during infection, which would have
been an irreparable confounder for our analysis. The difference is clear if
PCA is performed on all sequences from a ``compartimentalized'' patient, p7.
This is shown in \figurename~\ref{fig:aftp}: late sequences subdivide in two
well-defined clusters -- this characteristic is absent from all the patient
sequences used in our main analysis. On the right, allele frequency trajectories from the same
patient are plotted: evidently, no derived allele is fixed after 1000 days
from seroconversion, but a lot of mutations swarm to frequencies around 0.5-0.8. Of
course, the fixation probability of new mutations in such a sample is very low,
but this is an effect of population structure rather than any
of the processes discussed in the main text (time-dependent selection, competition
of mutants within the same epitope).

### 2013-02-04 synmut.tex
all experimental scripts (but no alignments yet, nor remote modules)
PCA of good patient in suppl

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fix_loss_dt_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations.
Panel A) Time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the  fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in C2-V5 is below the neutral
expectation indicated by the diagonal line. This suppression is not
oberserved in other parts of {\it env} or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

@@

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the deleterious
effect of synonymous mutations, the fraction of synonymous mutations that are
deleterious, the fitness effect of adaptive mutations (escape rate) and the frequency of
escapes. Among all simulations, we select the ones that show $A_\text{syn}$ and
$P_\text{interm}$ as observed in the data, i.e., a large depression in fixation
probability of synonymous mutations but, simulteneously, a moderately high
synonymous diversity. Specifically, \FIG{simsfig} shows parameter combinations
for which we found $A_\text{syn} < -0.15$ and $0.0025 < P_\text{interm} <
0.010$. This selection imposes a strong
constraint on the deleterious fraction and effect size (see \FIG{simsfig}).
A high fraction ($\gtrsim 0.8$) of only slightly deleterious synonymous sites,
with effect size $|s_d| \sim 0.002$, is required. This result fits well the
expectation based on the fixation/extinction times above (see \FIG{fixp1}).
The results are plausible:
(a) since high frequency bins are enriched in neutral mutations (it is easier
for them to hitchhike), a depression in $\pfix$ will be
visible only if deleterious mutations are pervasive; (b) in order to
hitchhike, the deleterious effect size has to be much smaller than the escape
rate, otherwise the
double mutant has little or no fitness advantage; (c) mutations with a 
deleterious effect weaker than approximately $0.001$ behave neutrally
consistent with the typical coalescent times observed in HIV.

>>>>

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations.
Panel A) Time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the  fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in C2-V5 is below the neutral
expectation indicated by the diagonal line. This suppression is not
oberserved in other parts of {\it env} or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

@@

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the deleterious
effect of synonymous mutations, the fraction of synonymous mutations that are
deleterious, the escape rate and the frequency of new escapes. Note that the
escape rate is sum of two factors: (a) the beneficial effect due to the ability to
evade the immune system minus (b) the fitness side costs in terms of structure,
stability, etc.
Among all simulations, we select the ones that show $A_\text{syn}$ and
$P_\text{interm}$ as observed in the data, i.e., a large depression in fixation
probability of synonymous mutations but, simulteneously, a moderately high
synonymous diversity. Specifically, \FIG{simsfig} shows parameter combinations
for which we found $A_\text{syn} < -0.15$ and $0.0025 < P_\text{interm} <
0.010$. This selection imposes a strong
constraint on the deleterious fraction and effect size (see \FIG{simsfig}).
A high fraction ($\gtrsim 0.8$) of only slightly deleterious synonymous sites,
with effect size $|s_d| \sim 0.002$, is required. This result fits well the
expectation based on the fixation/extinction times above (see \FIG{fixp1}).
The results are plausible:
(a) since high frequency bins are enriched in neutral mutations (it is easier
for them to hitchhike), a depression in $\pfix$ will be
visible only if deleterious mutations are pervasive; (b) in order to
hitchhike, the deleterious effect size has to be much smaller than the escape
rate, otherwise the
double mutant has little or no fitness advantage; (c) mutations with a 
deleterious effect weaker than approximately $0.001$ behave neutrally
consistent with the typical coalescent times observed in HIV.
